XSGRB - The X-shooter GRB afterglow sample

### 2017-01-09 document/XSGRB_sample.tex
Beginning introduction

\abstract{
The {\it Swift} satellite allows us to use gamma-ray bursts (GRBs) to peer into 
the hearts of star forming galaxies through cosmic time. Our open
collaboration, representing most of the active ESO member researchers in this
field, seeks to build a public legacy sample of GRB X-shooter spectroscopy
while {\it Swift} continues to fly. We propose to continue our programme to target
all suitably observable GRB afterglows (up to 15 bursts per semester), with the
primary goal of producing a well-defined, homogeneous, statistically useful
sample. To date, our spectroscopy covers a redshift range from 0.059 to about
8, with more than 20 robust metallicity measurements from absorption lines (over 
the redshift range 1.7--5.9) and 4 secure detections of H$_2$ or CH molecular
absorption. Such information is extremely difficult to obtain by other means.
In terms of studying the spread and redshift evolution in gas-phase properties, the
sample is still limited by low-number statistics. 
}

@@

Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ QSOs have 5 systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified
\citep{Noterdaeme2012b}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep{Jakobsson2006a, Fynbo2009}. Whereas DLAs
towards QSOs are mostly limited to $1.8\lesssim z \lesssim 5$ due to the
atmospheric  UV-cutoff and increasing Lyman-blanketing at increasing redshifts
\citep[e.g.,][]{Rafelski2014}, GRBs allow us to see into the hearts of
star-forming galaxies over the full history of cosmic star formation from $z
\approx 0$ to $z > 8$ \citep[e.g.,][]{Tanvir2009, Salvaterra2009,
	Jakobsson2012}. With afterglow spectroscopy (throughout the electromagnetic
spectrum from X-rays to the sub-mm) we can hence characterize the properties of
star-forming galaxies over cosmic history in terms of  redshifts, metallicities,
molecular contents, ISM temperatures, UV-flux densities, etc.. This is, however,
only possible as long as there are satellites in orbit that rapidly and
accurately locate GRBs. The currently operating \textit{Swift} satellite,
launched in 2004 and still fully-functioning, allows for very efficient
follow-up observations of GRBs due to its unprecedented rate, speed and
precision of localisations.

There are several unanswered but fundamental questions that must be addressed in
order to exploit the full potential of GRBs as cosmological probes. More than
50\% of the \textit{Swift} bursts with measured redshift are at $z > 2$, and
5--10\% are expected to be above $z=5$ \citep{Salvaterra2007,
	Salvaterra2012,Jakobsson2012, Perley2016a}. A high redshift completeness is
crucial for our understanding of the link between the number density of GRBs per
unit redshift and the global star-formation history of the Universe, as measured
by other means \citep[UV, FIR, sub-mm, see][]{Robertson2012}. The detection of
GRBs at $z > 6$ shows that GRBs have become competitive as a tool to identifying
galaxies at the highest redshifts and unsurpassed in providing detailed
abundance information via absorption line spectroscopy \citep{Tanvir2012,
	McGuire2016}.

\smallskip

%The sample of short GRBs with redshifts remains too small and too
%incomplete to really
%constrain their redshift distribution (Guetta \& Piran 2006, A\&A, 453, 823;
%Salvaterra et al.\ 2008, MNRAS, 388, 6) making it of extreme importance to
%enlarge as far as possible the number of known redshifts for this class of
%sources.

From March 2005 to March 2016 there have been about 350 {\it Swift} bursts
complying with our sample selection criteria (see ``Immediate objective''), and
about half of them have measured redshifts. Among the latter subset, the team
proposing these observations has measured about two thirds of the redshifts,
mainly with FORS1/2 and X-shooter \citep[see][]{Fynbo2009, Jakobsson2012, Kruhler2012}(see Fynbo et al.\ 2009, ApJS, 185, 526; 
Jakobsson et al.\ 2012, ApJ, 752, 62; Kr\"uhler et al.\ 2012, ApJ, 758, 46). Our
current aim is to build a sample superior to our previous low-resolution survey
(Fynbo et al.\ 2009, ApJS, 185, 526), both in terms of quantity and  quality of
the spectra. Our program started as guaranteed time observations during periods
84-91, and we have continued in open time since then.

%In P84 we started the X-shooter legacy survey in GTO time. Our aim is to  build
%a sample of GRBs superior to our previous  low-resolution  survey, Fynbo et
% al.\
%2009, ApJS, 185, 526) both in terms of quantity and  quality of the spectra.
% The
%last period of GTO time on the survey was P90 (plus 6 hr of leftover GTO time
%in  P91) and for P92 a proposal like this was  accepted for continuation of the
%program in open time (concerning P93 see box 8).

\smallskip

X-shooter is in many ways the ideal GRB follow-up instrument and indeed GRB
follow-up was one of the primary science cases behind the instrument design and
implementation. Our program secures general purpose GRB afterglow spectroscopic
follow-up that adds strong legacy value to the \textit{Swift} GRB sample. Due to
the wide wavelength range of X-shooter with the same observation cover molecular
H$_2$ absorption near the atmospheric cut-off and all the strong emission lines
from the host in the NIR arm (e.g., Friis et al., 2015, MNRAS, 451, 167). In
general, the wide wavelength coverage ensures that we always have features on
which to base a redshift measurement as long as the afterglow is brighter than
about 23 mag in either the $R$- or $z$-band. Frequently, emission lines are also
detected from the underlying host, which also provide further information such
as SFR and metallicity (the top right panel in Fig.~1 shows an example). Only
for 7 out of more than 70 secured spectra could we not measure a redshift. With
the X-shooter survey we provide {\bf metallicity measurements} for about 30\%
(Voigt-profile fits) of the $z>1.7$ events. So far we have measured
metallicities for more than 20 GRB afterglows with X-shooter. With the wide
wavelength coverage of X-shooter we can study important chemical species as
Zn\,\textsc{ii}, Cr\,\textsc{ii} and $\alpha$ elements over a much wider
redshift range than what is possible with other instruments. As an example, we
have measured a metallicity of $0.1 Z_\odot$ for GRB\,100219A at $z=4.669$
(Th{\"o}ne et al.\ 2013, MNRAS, 428, 3590), $0.02Z_\odot$ for GRB\,111008A at
$z=4.991$ (Sparre et al.\ 2014, ApJ, 785, 150) and $0.05Z_\odot$ for the
$z=5.9125$ GRB\,130606A (Hartoog et al.\ 2015, A\&A, 580, 139). Reconciling the
abundance patterns of GRB absorbers, other types of absorbers, QSO DLAs in
particular, and old stars in the Local Group is an important long-term goal (see
also Sparre et al.\ 2014, ApJ, 785, 150). Metallicities are also measured from
host emission lines (Kr{\"u}hler et al.\ 2015, A\&A, 581, A125). GRB
spectroscopy also allows us to determine the dust content of their environments,
both through analysis of the depletion pattern and through measurement of the
associated extinction (Japelj et al.\ 2015, A\&A, 451, 2050). This allows us to
quantify the dust-to-metals ratio and its evolution with redshift (e.g., De Cia
et al.\ 2013, A\&A, 560, 88; Zafar \& Watson 2013, A\&A, 560, 26). %Moreover,
% GRB
%afterglows allow investigation of random foreground absorbers, complementing
%QSO sightline studies (e.g., Schulze et
%al.\ 2012, A\&A, 546, 20).

%In Fynbo et al. (2009, ApJS, 185,
%526), we compare the host-galaxy HI
%column density and metallicity (most of which has been measured by our group)
%with a sample of damped \lya\ systems (DLAs) along QSO sightlines. We show that
%most GRB-DLAs have a much larger column than QSO-DLAs.  The difference in
%column density and metallicity for GRB and QSO sightlines shows that the total
%cross-section for environments like those of GRBs must be less than a few
%percent of the total cross-section for QSO-DLAs, and suggests that GRB
%sightlines probe regions with high gas density.

%The issue of production and {\bf release of Lyman continuum radiation} is
%currently not fully understood, but the evidence suggests that star-forming
%galaxies are dominating, in particular at the highest redshifts (e.g., Bianchi
%et al.\ 2001, A\&A, 376, 1; Chen et al.\ 2007, ApJ, 667, L25). Also here
%X-shooter spectroscopy will allow major progress due to the excellent blue
%sensitivity.

%GRB afterglows can probe dust in high-redshift star-forming galaxies
%including exploring the {\bf 2175~\AA{} extinction bump}
%(El\'iasd\'ottir et al.\ 2009, ApJ, 697, 1725; Zafar et al.\ 2011, A\&A,
%732, 143).
%Due to the broad wavelength  coverage we can do
%this more accurately and over a broader range of redshifts with X-shooter.

\smallskip

We will also determine the frequency and properties of {\bf molecular
	absorption} towards GRB absorbers. Molecular gas is a key element to catalyze
the process of star formation, but prior to our program H$_2$ had been detected
just in two cases (tentatively in Fynbo et al.\ 2006, A\&A, 451, L47; securely
in Prochaska et al.\ 2009, ApJ, 691, L27). With our X-shooter program we have
found three more systems (Kr\"uhler et al.\ 2013, A\&A, 557, 18; D'Elia et al.\
2014, A\&A, 564, 38; Friis et al.\, 2015, MNRAS, 451, 167). We are currently
analysing more of our spectra for less obvious molecular absorption and we
expect to find more (a dedicated sample paper is addressing this issue).

\vspace{0.2cm} A natural question to ask is: {\bf how long should this work
	continue?} Our view is that we need to keep observing the afterglows as long as
we have \textit{Swift} in operation. Also note that the program is still
producing many papers and provides data for many theses (Box 9 and 10).
\textit{Swift} is currently funded until 2018, but is likely to get more
extensions given its overwhelming success. As mentioned above GRBs allow us to
probe star-forming galaxies that are almost impossible to study in other ways
both in terms of redshifts, galaxy luminosity function, and regions within
galaxies. After 7 periods we have secured seven spectra of $z>4$ GRBs, of which
three were of sufficient quality to allow abundance measurements (Th\"one et
al.\ 2013, Sparre et al.\ 2014, Hartoog et al.\ 2015). GRBs offer the only way
to derive chemical abundances for the gas phase of central, actively
star-forming regions of high-$z$ galaxies. The program also maintains a very
high discovery potential where we occasionally find something completely
unexpected that provides interesting clues to puzzles in other fields, e.g.
extinction of type Ia supernovae (Fynbo et al.\ 2014, A\&A, 572, 12). Each of
these spectra are like precious jewels – it is a type of observation that can
never be repeated and a class of sightlines that can only be studied while we
have operating GRB satellites.

It is also worth adding that we have build up a rather unique team spread over
Europe from Granada to Reykjavik, which by now has reached a point where the
distribution of night shifts, the scientific exploitation of the data is
efficient and where we are open to all new members who wish to participate. As
mentioned all data are public immediately.

{\bf For all of these reasons, we need to keep building up the sample of GRB
	afterglow spectra now as we may have to wait many years before a mission like
	\textit{Swift} becomes available again.}

A significant proportion of GRBs lack a bright optical afterglow (``{\bf dark
	bursts}'', e.g., Jakobsson et al.\ 2004, ApJ, 617, L21; Melandri et al.\ 2012,
MNRAS, 421, 1265). Some of these are at the highest redshifts ($z > 6$) and
their observer-frame optical emission is absorbed by the IGM. The majority,
however, suffer from large dust obscuration (e.g., Perley et al. 2009, AJ, 138,
1690; Greiner et al.\ 2011, A\&A, 526, 30). Identifying such GRBs is important
for constraining the fraction of obscured star formation. In both cases, NIR
emission is expected. X-shooter can adequately study these objects, provided
that a NIR counterpart is timely identified, for which we have the dedicated
HAWKI run D.

The detection of {\bf absorption line variability} can reveal the burst
influence on the surrounding medium and in turn the absorber distance from the
burst and its metallicity (Vreeswijk et al.\ 2007, A\&A 468, 83; D'Elia et al.\
2009, ApJ, 694, 332; Th\"one et al.\ 2011, MNRAS, 414, 479; De Cia et al.\ 2012,
A\&A, 545, 64; Hartoog et al.\ 2013). {\bf Short GRBs} originate in a
substantially different environment compared to long GRBs. Short GRBs may be
related to the merging of compact binaries and the coalescence time can be long
enough to allow the progenitor system to move far away from the star formation
site (Belczynski et al.\ 2002, ApJ, 571, 147). Up to now, however, no spectrum
with a sufficient signal-to-noise ratio of a short GRB afterglow has been
secured. A knowledge of the redshift distribution of short bursts is of key
importance for the next generation of gravitational wave experiments, as they
are the likely EM counterparts to their primary targets.

@@

Because we only follow up targets with either a clear afterglow or counterpart,
correctly associating a galaxy with a burst when there is no detected optical afterglow
is more difficult, meaning that we should not have attempted follow-up for a
larger fraction of bursts with no optical afterglow. The X-ray properties for
the followed sample does not differ signicantly from the statistical sample, as
shown in Sect. \ref{completeness}. Using the table maintained by
Greiner\footnote{http://www.mpe.mpg.de/~jcg/grbgen.html} \todo{How to reference
this}, we can see how the pressence of an optical afterglow affects the
follow-up statistics. 50.5 percent of all \textit{Swift}-detected bursts do not
have an detected optical afterglow, but this number also includes bursts where
no optical telescopes were available, so is a lower limit. For the bursts
entering our statistical sample, this number is 27 percent

Since \textit{all} GRBs should have an optical afterglow, the presence of an optical afterglow is more a statement about the sensitivity than optical properties of the bursts.

>>>>

\abstract{
In this work we present the spectra of \textit{all} $\gamma$-ray bursts (GRBs)
afterglows that have \textit{ever} been observed at the VLT/X-shooter. This
totals 137 spectroscopic observation of 118 individual bursts, of which
redshifts have been secured for 91 \% covering a redshift range from 0.059 to
$\sim$ 8.5. Based on a set of unbiasing selection criteria, the follow-up effort
has been guided by the \textit{Swift}-telescope to produce a homogeneous sample
of afterglows for 86 spectra, representing the underlying \textit{Swift} GRB population.
We here provide a public release of all the spectra, including continuum
estimates and telluric corrections. We provide an assessment of the degree of
completeness with respect to the parent GRB population, in terms of the X-ray
properties of the bursts in the sample and find that the sample presented here
can not be significantly distinguished from the \textit{Swift} sample.
For all bursts (39) for which it is possible, we provide a measurement of the
optically derived hydrogen column density, increasing the total number of
published HI column densities by $\sim$ 50 \% and characterize the distribution
compared with literature values.
We constrain the fraction of dark burst to be 27\% - 42\% and confirm previous
results that higher optical darkness has increased X-ray excess
absorption.
%This dataset provides a unique possibility to study the ISM across cosmic time.
}

@@

For a brief moment, the energy output of a $\gamma$-ray burst (GRB) can rival
that of the stellar light integrated over the entirety of the observable
universe \citep{Kumar2015}. Bright across the entire electromagnetic spectrum,
the following afterglow makes possible the exploration of the intervening space,
otherwise dark. Since the discovery of the first $\gamma$-ray counterparts
nearly two decades ago\citep{Costa1997a, VanParadijs1997, Djorgovski1997,
	Frail1997}, significant advances in the understanding of the underlying
phenomenology and potential usage has been made.

Categorized on the basis of their duration and spectral hardness, two main classes of GRBs exist, long and short \citep{Kouveliotou1993}. Now thoroughly established, the connection between long GRBs and supernovae (SNe) \citep{Galama1998, Hjorth2003a, Stanek2003, Woosley2006c} and the association of the location of long GRBs with the most star-forming regions in their host galaxies \citep{Fruchter2006, Svensson2010a}, make a stellar origin a likely explanation for the progenitor system. The seeming metal aversion \citep{Graham2013, Kruhler2015, Perley2016b} in selection of host at low redshift along with theoretical single-progenitor models preferring more pristine starting conditions \citep{Woosley2006d}, paints a picture where the core of a rapidly spinning, massive star undergoes gravitational collapse, although a certain degree of binarity \citep{Fryer2005, Podsiadlowski2010} in the progenitor population is required to explain the host metallicity distribution to super solar values \citep{Niino2011}. The driving engine in short GRBs have been established as the merger of two compact objects \citep{Metzger2010} through the detection of an associated kilonova \citep{Tanvir2013, Berger2013a}.

 see \citet{Levan2016} for recent review.

%Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
%towards about 10$^5$ QSOs have 5 systems with
%$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified
%\citep{Noterdaeme2012b}. Long GRB afterglow spectra, by contrast, reveal such
%systems in the majority of cases \citep{Jakobsson2006a, Fynbo2009}. Whereas DLAs
%towards QSOs are mostly limited to $1.8\lesssim z \lesssim 5$ due to the
%atmospheric  UV-cutoff and increasing Lyman-blanketing at increasing redshifts
%\citep[e.g.,][]{Rafelski2014}, GRBs allow us to see into the hearts of
%star-forming galaxies over the full history of cosmic star formation from $z
%\approx 0$ to $z > 8$ \citep[e.g.,][]{Tanvir2009, Salvaterra2009,
%	Jakobsson2012}. With afterglow spectroscopy (throughout the electromagnetic
%spectrum from X-rays to the sub-mm) we can hence characterize the properties of
%star-forming galaxies over cosmic history in terms of  redshifts, metallicities,
%molecular contents, ISM temperatures, UV-flux densities, etc.. 
%
%
%This is, however,
%only possible as long as there are satellites in orbit that rapidly and
%accurately locate GRBs. The currently operating \textit{Swift} satellite,
%launched in 2004 and still fully-functioning, allows for very efficient
%follow-up observations of GRBs due to its unprecedented rate, speed and
%precision of localisations.
%
%There are several unanswered but fundamental questions that must be addressed in
%order to exploit the full potential of GRBs as cosmological probes. More than
%50\% of the \textit{Swift} bursts with measured redshift are at $z > 2$, and
%5--10\% are expected to be above $z=5$ \citep{Salvaterra2007,
%	Salvaterra2012,Jakobsson2012, Perley2016a}. A high redshift completeness is
%crucial for our understanding of the link between the number density of GRBs per
%unit redshift and the global star-formation history of the Universe, as measured
%by other means \citep[UV, FIR, sub-mm, see][]{Robertson2012}. The detection of
%GRBs at $z > 6$ shows that GRBs have become competitive as a tool to identifying
%galaxies at the highest redshifts and unsurpassed in providing detailed
%abundance information via absorption line spectroscopy \citep{Tanvir2012,
%	McGuire2016}.
%
%\smallskip
%
%%The sample of short GRBs with redshifts remains too small and too
%%incomplete to really
%%constrain their redshift distribution (Guetta \& Piran 2006, A\&A, 453, 823;
%%Salvaterra et al.\ 2008, MNRAS, 388, 6) making it of extreme importance to
%%enlarge as far as possible the number of known redshifts for this class of
%%sources.
%
%From March 2005 to March 2016 there have been about 350 {\it Swift} bursts
%complying with our sample selection criteria, and
%about half of them have measured redshifts. Among the latter subset, the team
%proposing these observations has measured about two thirds of the redshifts,
%mainly with FORS1/2 and X-shooter \citep[see][]{Fynbo2009, Jakobsson2012, Kruhler2012}. Our
%current aim is to build a sample superior to our previous low-resolution survey
%\citep{Fynbo2009}, both in terms of quantity and  quality of
%the spectra. 
%
%Due to
%the wide wavelength range of X-shooter with the same observation cover molecular
%H$_2$ absorption near the atmospheric cut-off and all the strong emission lines
%from the host in the NIR arm \citep[e.g.,][]{Friis2015}. In
%general, the wide wavelength coverage ensures that we always have features on
%which to base a redshift measurement as long as the afterglow is brighter than
%about 23 mag in either the $R$- or $z$-band. Frequently, emission lines are also
%detected from the underlying host, which also provide further information such
%as SFR and metallicity (the top right panel in Fig.~1 shows an example). Only
%for 7 out of more than 70 secured spectra could we not measure a redshift. With
%the X-shooter survey we provide {\bf metallicity measurements} for about 30\%
%(Voigt-profile fits) of the $z>1.7$ events. So far we have measured
%metallicities for more than 20 GRB afterglows with X-shooter. With the wide
%wavelength coverage of X-shooter we can study important chemical species as
%Zn\,\textsc{ii}, Cr\,\textsc{ii} and $\alpha$ elements over a much wider
%redshift range than what is possible with other instruments. As an example, we
%have measured a metallicity of $0.1 Z_\odot$ for GRB\,100219A at $z=4.669$
%(Th{\"o}ne et al.\ 2013, MNRAS, 428, 3590), $0.02Z_\odot$ for GRB\,111008A at
%$z=4.991$ (Sparre et al.\ 2014, ApJ, 785, 150) and $0.05Z_\odot$ for the
%$z=5.9125$ GRB\,130606A (Hartoog et al.\ 2015, A\&A, 580, 139). Reconciling the
%abundance patterns of GRB absorbers, other types of absorbers, QSO DLAs in
%particular, and old stars in the Local Group is an important long-term goal (see
%also Sparre et al.\ 2014, ApJ, 785, 150). Metallicities are also measured from
%host emission lines (Kr{\"u}hler et al.\ 2015, A\&A, 581, A125). GRB
%spectroscopy also allows us to determine the dust content of their environments,
%both through analysis of the depletion pattern and through measurement of the
%associated extinction (Japelj et al.\ 2015, A\&A, 451, 2050). This allows us to
%quantify the dust-to-metals ratio and its evolution with redshift (e.g., De Cia
%et al.\ 2013, A\&A, 560, 88; Zafar \& Watson 2013, A\&A, 560, 26). %Moreover,
%% GRB
%%afterglows allow investigation of random foreground absorbers, complementing
%%QSO sightline studies (e.g., Schulze et
%%al.\ 2012, A\&A, 546, 20).
%
%%In Fynbo et al. (2009, ApJS, 185,
%%526), we compare the host-galaxy HI
%%column density and metallicity (most of which has been measured by our group)
%%with a sample of damped \lya\ systems (DLAs) along QSO sightlines. We show that
%%most GRB-DLAs have a much larger column than QSO-DLAs.  The difference in
%%column density and metallicity for GRB and QSO sightlines shows that the total
%%cross-section for environments like those of GRBs must be less than a few
%%percent of the total cross-section for QSO-DLAs, and suggests that GRB
%%sightlines probe regions with high gas density.
%
%%The issue of production and {\bf release of Lyman continuum radiation} is
%%currently not fully understood, but the evidence suggests that star-forming
%%galaxies are dominating, in particular at the highest redshifts (e.g., Bianchi
%%et al.\ 2001, A\&A, 376, 1; Chen et al.\ 2007, ApJ, 667, L25). Also here
%%X-shooter spectroscopy will allow major progress due to the excellent blue
%%sensitivity.
%
%%GRB afterglows can probe dust in high-redshift star-forming galaxies
%%including exploring the {\bf 2175~\AA{} extinction bump}
%%(El\'iasd\'ottir et al.\ 2009, ApJ, 697, 1725; Zafar et al.\ 2011, A\&A,
%%732, 143).
%%Due to the broad wavelength  coverage we can do
%%this more accurately and over a broader range of redshifts with X-shooter.
%
%\smallskip
%
%We will also determine the frequency and properties of {\bf molecular
%	absorption} towards GRB absorbers. Molecular gas is a key element to catalyze
%the process of star formation, but prior to our program H$_2$ had been detected
%just in two cases (tentatively in Fynbo et al.\ 2006, A\&A, 451, L47; securely
%in Prochaska et al.\ 2009, ApJ, 691, L27). With our X-shooter program we have
%found three more systems (Kr\"uhler et al.\ 2013, A\&A, 557, 18; D'Elia et al.\
%2014, A\&A, 564, 38; Friis et al.\, 2015, MNRAS, 451, 167). We are currently
%analysing more of our spectra for less obvious molecular absorption and we
%expect to find more (a dedicated sample paper is addressing this issue).
%
%\vspace{0.2cm} A natural question to ask is: {\bf how long should this work
%	continue?} Our view is that we need to keep observing the afterglows as long as
%we have \textit{Swift} in operation. Also note that the program is still
%producing many papers and provides data for many theses (Box 9 and 10).
%\textit{Swift} is currently funded until 2018, but is likely to get more
%extensions given its overwhelming success. As mentioned above GRBs allow us to
%probe star-forming galaxies that are almost impossible to study in other ways
%both in terms of redshifts, galaxy luminosity function, and regions within
%galaxies. After 7 periods we have secured seven spectra of $z>4$ GRBs, of which
%three were of sufficient quality to allow abundance measurements (Th\"one et
%al.\ 2013, Sparre et al.\ 2014, Hartoog et al.\ 2015). GRBs offer the only way
%to derive chemical abundances for the gas phase of central, actively
%star-forming regions of high-$z$ galaxies. The program also maintains a very
%high discovery potential where we occasionally find something completely
%unexpected that provides interesting clues to puzzles in other fields, e.g.
%extinction of type Ia supernovae (Fynbo et al.\ 2014, A\&A, 572, 12). Each of
%these spectra are like precious jewels – it is a type of observation that can
%never be repeated and a class of sightlines that can only be studied while we
%have operating GRB satellites.
%
%It is also worth adding that we have build up a rather unique team spread over
%Europe from Granada to Reykjavik, which by now has reached a point where the
%distribution of night shifts, the scientific exploitation of the data is
%efficient and where we are open to all new members who wish to participate. As
%mentioned all data are public immediately.
%
%{\bf For all of these reasons, we need to keep building up the sample of GRB
%	afterglow spectra now as we may have to wait many years before a mission like
%	\textit{Swift} becomes available again.}
%
%A significant proportion of GRBs lack a bright optical afterglow (``{\bf dark
%	bursts}'', e.g., Jakobsson et al.\ 2004, ApJ, 617, L21; Melandri et al.\ 2012,
%MNRAS, 421, 1265). Some of these are at the highest redshifts ($z > 6$) and
%their observer-frame optical emission is absorbed by the IGM. The majority,
%however, suffer from large dust obscuration (e.g., Perley et al. 2009, AJ, 138,
%1690; Greiner et al.\ 2011, A\&A, 526, 30). Identifying such GRBs is important
%for constraining the fraction of obscured star formation. In both cases, NIR
%emission is expected. X-shooter can adequately study these objects, provided
%that a NIR counterpart is timely identified, for which we have the dedicated
%HAWKI run D.
%
%The detection of {\bf absorption line variability} can reveal the burst
%influence on the surrounding medium and in turn the absorber distance from the
%burst and its metallicity (Vreeswijk et al.\ 2007, A\&A 468, 83; D'Elia et al.\
%2009, ApJ, 694, 332; Th\"one et al.\ 2011, MNRAS, 414, 479; De Cia et al.\ 2012,
%A\&A, 545, 64; Hartoog et al.\ 2013). {\bf Short GRBs} originate in a
%substantially different environment compared to long GRBs. Short GRBs may be
%related to the merging of compact binaries and the coalescence time can be long
%enough to allow the progenitor system to move far away from the star formation
%site (Belczynski et al.\ 2002, ApJ, 571, 147). Up to now, however, no spectrum
%with a sufficient signal-to-noise ratio of a short GRB afterglow has been
%secured. A knowledge of the redshift distribution of short bursts is of key
%importance for the next generation of gravitational wave experiments, as they
%are the likely EM counterparts to their primary targets.

@@

Because we only follow up targets with either a clear afterglow or counterpart,
correctly associating a galaxy with a burst when there is no detected optical
afterglow is more difficult, see \citet{Levesque2010} and \citet{Perley2017} for
a recent example. Because the presence of optical transient is highly dependent
on the "darkness", we expect that the $\beta_{OX}$ sample is biased against dark
bursts. The larger relative fraction of upper limits for increased darkness is
visible both in the sample presented here and the one in \citet{Fynbo2009}. The
X-ray properties for the followed sample does not differ signicantly from the
statistical sample, as shown in Sect. \ref{completeness}. Using the table
maintained by Greiner\footnote{http://www.mpe.mpg.de/~jcg/grbgen.html} \todo{How
to reference this}, we can see how the presence of an optical afterglow affects
the follow-up statistics. 50.5 percent of all \textit{Swift}-detected bursts do
not have an detected optical afterglow, but this number also includes bursts
where no optical telescopes were available, so is a lower limit. For the bursts
entering our statistical sample, this number is $\sim$ 27 percent, close to the
maximum fraction of dark bursts \citep{Melandri2012}. Of the subsample (3) that
have actually been attempted, this number is $\sim$ 18 percent, confirming that
we are indeed biased against bursts with an optical afterglow in the
spectroscopic sample.

Since \textit{all} GRBs should have an optical afterglow, the presence of an optical afterglow is more a statement about the sensitivity than optical properties of the bursts. \todo{Daniele, help.}
